# Spatial distribution models


This page shows how you can use the Random Forest algorithm to make spatial predictions. This approach is widely used, for example to classify remote sensing data into different land cover classes. But here our objective is to predict the entire range of a species based on a set of locations where it has been observed. As an example, we use the hominid species *Imaginus magnapedum* (also known under the vernacular names of "bigfoot" and "sasquatch"). This species is believed to occur in the United States, but it is so hard to find by scientists that its very existence is commonly denied by the mainstream media --- despite the many reports on Twitter! For more information about this controversy, see the article by Lozier, Aniello and Hickerson: [Predicting the distribution of Sasquatch in western North America: anything goes with ecological niche modelling](http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2699.2009.02152.x/abstract).

We will use "citizen-science" data to find out: 
   
   a) What the complete range of the species might be.

   b) How good (general) our model is by predicting the range of the Eastern sub-species, with data from the Western sub-species. 
   
   c) How climate change might affect its distribution.
   
In this context, this type of analysis is often referred to as 'species distribution modeling' or 'ecological niche modeling'. [Here is a more in-depth discussion](/terra/sdm) of this technique.

First make sure we have the packages needed:

```{r pkgs, echo=TRUE}
if (!require("rspat")) remotes::install_github("rspatial/rspat")
if (!require("predicts")) remotes::install_github("rspatial/predicts")
if (!require("geodata")) install.packages("geodata")
```  
  
## Data 

### Observations

We get a data set of reported Bigfoot observations

```{r getData, echo=TRUE}
library(rspat)
bf <- spat_data("bigfoot")
dim(bf)
head(bf)
```

It is always good to first plot the locations to see what we are dealing with.

```{r, sup1}
plot(bf[,1:2], cex=0.5, col="red")

library(geodata)
wrld <- geodata::world(path=".")
bnds <- wrld[wrld$NAME_0 %in% c("Canada", "Mexico", "United States"), ]
lines(bnds)
```

So the are in Canada and in the United States, but no reports from Mexico, so far.

### Predictor variables

Here, as is common in species distribution modeling, we use climate data as predictor variables in our model. Specifically, we use "bioclimatic variables", see: https://www.worldclim.org/data/bioclim.html. Here we used a spatial resolution of 10 minutes (one sixt of a degree). That is relatively coarse but it makes the download and processing faster.

```{r, sup3}
wc <- geodata::worldclim_global("bio", res=10, ".")
plot(wc[[c(1, 12)]], nr=2)
```

Now extract climate data for the locations of our observations. In that way, we can find out what the climate conditions are that the species likes, apparently.

```{r, sup5x}
bfc <- extract(wc, bf[,1:2])
head(bfc, 3)
```

I remove the first column with the ID that we do not need. 

```{r, sup5}
bfc <- bfc[,-1]
```


Now we can plot the species' distribution in a part of the *environmental* space. Here is a plot of temperature vs rainfall of sites where Bigfoot was observed. 

```{r, sup7}
plot(bfc[ ,"wc2.1_10m_bio_1"], bfc[, "wc2.1_10m_bio_12"], col="red",
        xlab="Annual mean temperature (°C)", ylab="Annual precipitation (mm)")
```


### Background data

Normally, one would build a model that would compare the values of the predictor variables at the locations where something was observed, with those values at the locations where it was not observed. But we do not have data from a systematic survey that determined presence and absence. We have presence-only data. (And, determining absence is not that simple. You blink and Bigfoot is gone!). 

The common approach to deal with these type of data is to not model presence versus absence, but presence versus "background". The "background" is the random (or maximum entropy) expectation; it is what you would get if the species had no preference for any of the predictor variables (or to other variables that are not in the model, but correlated with the predictor variables).

There is not much point in taking absence data from very far away (tropical Africa or Antarctica). Typically they are taken from more or less the entire study area for which we have presences data.

To do so, I first get the extent of all points
```{r}
ext_bf <- ext(vect(bf[, 1:2])) + 1
ext_bf
```

And then I take 5000 random samples (excluding `NA` cells) from SpatExtent `e`, by using it as a "window" (blacking out all other areas) on the climate SpatRaster.

```{r}
set.seed(0)
window(wc) <- ext_bf
bg <- spatSample(wc, 5000, "random", na.rm=TRUE, xy=TRUE)
head(bg)
```

Instead of using `window` you could also subset the climate data like this `wc <- crop(wc, ext_bf)`

Above, with `spatSample`, I used the argument `xy=TRUE` to be able to show were these points are from:

```{r}
plot(bg[, c("x", "y")])
```

But we otherwise do not need them so I remove them again,

```{r}
bg <- bg[, -c(1:2)]
```


We can now compare the climate of the presence and background points, for example, for temperature and rainfall

```{r}
plot(bg[,1], bg[,12], xlab="Annual mean temperature (°C)", 
  	      ylab="Annual precipitation (mm)", cex=.8)

points(bfc[,1], bfc[,12], col="red", cex=.6, pch="+")
legend("topleft", c("observed", "background"), col=c("red", "black"), pch=c("+", "o"), pt.cex=c(.6, .8))   
```

So we see that while Bigfoot is widespread, it is not common in cold areas, nor in hot and dry areas.


### East vs West

I am first going to split the data into East and West. This is because I believe there are two sub-species (The Eastern Sasquatch is darker, less hairy, and has more pointy ears). I am principally interested in the western sub-species. Note how I use the original coordinates to subset the climate data. We can do this because they are in the same order. 

```{r}
#eastern points
bfe <- bfc[bf[,1] > -102, ]
#western points
bfw <- bfc[bf[,1] <= -102, ]
```

And now I combine the presence ("1") with the background ("0") data (I use the same background data for both subspecies) 

```{r}
dw <- rbind(cbind(pa=1, bfw), cbind(pa=0, bg))
de <- rbind(cbind(pa=1, bfe), cbind(pa=0, bg))

dw <- data.frame(dw)
de <- data.frame(na.omit(de))

dim(dw)
dim(de)
```

## Fit a model

Now we have the data to fit a model. Let's first look at a Classification and Regression Trees (CART) model.


### CART

```{r, sup10a}
library(rpart)
cart <- rpart(pa~., data=dw)
```

The "complexity parameter" can be used as a stopping parameter to avoid overfitting. 

```{r, sup10ab}
printcp(cart)
plotcp(cart)
```

Fit the model again, with fewer splits

```{r, sup10abc}
cart <- rpart(pa~., data=dw, cp=0.02)
```


And here is the tree

```{r, sup10b}
library(rpart.plot)
rpart.plot(cart, uniform=TRUE, main="Regression Tree")
```


__Question 1__: *Describe the environmental conditions that Bigfoot appears to enjoy most?*


And now we can use the model to show how attractive the climate is for this species.

```{r sup10bb, fig.width=6, fig.height=4}
x <- predict(wc, cart)
x <- mask(x, wc[[1]])
x <- round(x, 2)
plot(x, type="class", plg=list(x="bottomleft"))
```

Notice that there are six values, because the regression tree has six leaves.

### Random Forest

CART gives us a nice result to look at that can be easily interpreted (as you just illustrated with your answer to Question 1). But the approach suffers from high variance (meaning that the model tends to be over-fit, it is different each time a somewhat different datasets are used); and the quality of its predictions suffers from that. Random Forest does not have that problem as much. Above, with CART, we use regression, let's do both regression and classification here. 

But first I set some points aside for validation (normally we would do k-fold cross-validation, but we keep it simple here).

```{r}
set.seed(123)
i <- sample(nrow(dw), 0.2 * nrow(dw))
test <- dw[i,]
train <- dw[-i,]
```

First we do classification, by making a categorical variable for presence/background.

```{r, sup11}
fpa <- as.factor(train[, 'pa'])
```

Now fit the RandomForest model

```{r, sup12a}
library(randomForest)
crf <- randomForest(train[, 2:ncol(train)], fpa)
crf
```

The Out-Of-Bag error rate is very small. 

The variable importance plot shows which variables are most important in fitting the model. 
This is computed by randomizing each variable one by one, and then evaluating the decline in model prediction.

```{r, sup12b}
varImpPlot(crf)
```

Now we use regression, rather than classification.
First we tune a parameter.

```{r, sup14a}
trf <- tuneRF(train[, 2:ncol(train)], train[, "pa"])
trf
mt <- trf[which.min(trf[,2]), 1]
mt
```

__Question 2__: *What did tuneRF help us find? What does the values of mt represent?*


```{r, sup14b}
rrf <- randomForest(train[, 2:ncol(train)], train[, "pa"], mtry=mt, ntree=250)
rrf
plot(rrf)
```


__Question 3__: *What does `plot(rrf)` show us?*


## Predict

We can use the model to make predictions to any other place for which we have values for the predictor variables. Our climate data is global so we could find suitable areas for Bigfoot in Australia, but let's stick to North America for now.

### Regression

```{r sup17a, fig.width=6, fig.height=4}
rp <- predict(wc, rrf, na.rm=TRUE)
plot(rp)
```

Note that the regression predictions are well-behaved, in the sense that they are between 0 and 1. However, they are continuous within that range, and if you wanted presence/absence, you would need a threshold. To get the optimal threshold, you would normally have a hold out data set, but here I use the training data for simplicity. 

```{r, sup17b}
library(predicts)
eva <- pa_evaluate(predict(rrf, test[test$pa==1, ]), predict(rrf, test[test$pa==0, ]))
eva
```


We can make a [ROC plot](http://gim.unmc.edu/dxtests/roc2.htm)
```{r, sup18}
plot(eva, "ROC")
```

This suggests that the model is (very near) perfect in disinguising presence from background points. This is perhaps better illustrated with these plots:

```{r, sup18b}
par(mfrow=c(1,2))
plot(eva, "boxplot")
plot(eva, "density")
```


To get a good threshold to determine presence/absence and plot the prediction, we can use the "max specificity + sensitivity" threshold.
```{r, sup19}
tr <- eva@thresholds
tr
plot(rp > tr$max_spec_sens)
```


### Classification

We can also use the classification Random Forest model to make a prediction.

```{r sup20a, fig.width=6, fig.height=4}
rc <- predict(wc, crf, na.rm=TRUE)
plot(rc)
```

They are different because the classification used a threshold of 0.5, which is not necessarily appropriate.

You can get probabilities for the classes (in this case there are 2 classes, presence and absence, and I only plot presence) 

```{r sup20b, fig.width=6, fig.height=4}
rc2 <- predict(wc, crf, type="prob", na.rm=TRUE)
plot(rc2, 2)
```


## Extrapolation

Now, let's see if our model is general enough to predict the distribution of the Eastern species. 

```{r sup22}
eva2 <- pa_evaluate(predict(rrf, de[de$pa==1, ]), predict(rrf, de[de$pa==0, ]))
eva2
par(mfrow=c(1,2))
plot(eva2, "ROC")
plot(eva2, "boxplot")
```

By this measure, it is a *terrible* model -- as we already saw on the map. So our model is really good in predicting the range of the West, but it cannot extrapolate at all to the East. 


```{r, sup24, fig.width=6, fig.height=4}
plot(rc)
points(bf[,1:2], cex=.25)
```


__Question 4__: *Why would it be that the model does not extrapolate well?*

An important question in the biogeography of the Bigfoot would be if it can survive in other parts of the world (it has been spotted trying to get on commerical flights leaving North America). 

Let's see.

```{r sup26, fig.width=6, fig.height=4}
window(wc) <- NULL
pm <- predict(wc, rrf, na.rm=TRUE)
plot(pm)
lines(wrld)
```

__Question 5__: *What are some countries that should consider Bigfoot as a potential invasive species?*


## Climate change 

We can also estimate range shifts due to climate change. We can use the same model, but now extrapolate in time (and space).


```{r sup28, fig.width=6, fig.height=4}
fut <- cmip6_world("CNRM-CM6-1", "585", "2061-2080", var="bio", res=10, path=".")
names(fut)
names(wc)
names(fut) <- names(wc)
window(fut) <- ext_bf
pfut <- predict(fut, rrf, na.rm=TRUE)
plot(pfut)
```

__Question 6__: *Make a map to show where conditions are improving for western bigfoot, and where they are not. Is the species headed toward extinction?*


## Further reading

More on [Species distribution modeling with R](/terra/sdm).

