# Local statistics

## Introduction

This handout accompanies Chapter 8 in [O'Sullivan and Unwin (2010)](http://www.wiley.com/WileyCDA/WileyTitle/productCd-0470288574.html). 

## LISA


We compute some measures of local spatial autocorrelation. 

First get the Auckland data.

```{r, loca1, message=FALSE}
if (!require("rspat")) remotes::install_github("rspatial/rspat")
library(rspat)
auck <- spat_data("auctb")
```

Now compute the spatial weights. You can try other ways (e.g. `relation="touches"`)

```{r, locasp}
w <- relate(auck, relation="rook")
``` 

Compute the Getis *Gi*

```{r, loca2}
Gi <- autocor(auck$TB, w, "Gi")
head(Gi)
```

And make a map

```{r, loca3, fig.height=4}
grays <- rev(gray(seq(0,1,.2)))
auck$Gi <- Gi
plot(auck, "Gi", col=grays)
```

Now on to the `Gi*` by including the focal area in the computation.

```{r, loca4, fig.height=4}
#include "self"
diag(w) <- TRUE
auck$Gistar <- autocor(auck$TB, w, "Gi*")
plot(auck, "Gistar", main="Gi*", col=grays)
```

This looks very similar to the local average, which we compute below.

```{r, loca6, fig.height=4}
auck$loc_mean <- apply(w, 1, function(i) mean(auck$TB[i]))
plot(auck, "loc_mean", main="Local mean", col=grays)
```

The local Moran *Ii* has not been implemented in `terra`, but we can use `spdep` (a package that if focussed on these type of statistics and on spatial regression). 

```{r, loca8, message=FALSE, fig.height=4}
library(spdep)
sfauck <- sf::st_as_sf(auck)
wr <- poly2nb(sfauck, row.names=sfauck$Id, queen=FALSE)
lstw <- nb2listw(wr, style="B")
auck$Ii <- localmoran(auck$TB, lstw)
plot(auck, "Ii", main="Local Moran", col=grays)
```

In the above I followed the book by useing the raw count data. Howwever, it would be more approproate to use disease density instead. You can compute that like this:

```{r}
auck$TBdens <- 10000 * auck$TB / expanse(auck)
```

The 10000 is just to avoid very small numbers. 


## Geographically weighted regression

Here is an example of GWR with California precipitation data. 


```{r, loca10}
cts <- spat_data("counties")
p <- spat_data("precipitation")
head(p)

plot(cts)
points(p[,c("LONG", "LAT")], col="red", pch=20)
```

Compute annual average precipitation
```{r, loca11}
p$pan <- rowSums(p[,6:17])
```

Global regression model
```{r, loca12}
m <- lm(pan ~ ALT, data=p)
m
```

Create `sf` objects with a planar crs.

```{r, loca13}
alb <- "+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +datum=WGS84"
sp <- vect(p, c("LONG", "LAT"), crs="+proj=longlat +datum=NAD83")
sp <- terra::project(sp, alb)
spsf <- sf::st_as_sf(sp)
vctst <- terra::project(cts, alb)
ctst <- sf::st_as_sf(vctst)
```

Get the optimal bandwidth
```{r, loca14}
library( spgwr )
bw <- gwr.sel(pan ~ ALT, crds(sp), data=spsf)
bw
``` 

Create a regular set of points to estimate parameters for.
```{r, loca16}
r <- rast(vctst, res=10000)
r <- rasterize(vect(ctst), r)
newpts <- geom(as.points(r))[, c("x", "y")]
```


Run the `gwr` function
```{r, loca17}
g <- gwr(pan ~ ALT, crds(sp), data=spsf, bandwidth=bw, fit.points=newpts[, 1:2])
g
```

Link the results back to the raster.

```{r, loca18, fig.width=9}
slope <- intercept <- r
slope[!is.na(slope)] <- g$SDF$ALT 
intercept[!is.na(intercept)] <- g$SDF$"(Intercept)"
s <- c(intercept, slope)
names(s) <- c("intercept", "slope")
plot(s)
```

See [this page](http://rspatial.org/terra/analysis/6-local_regression.html) for a more detailed example of geographically weighted regression.
