# Fields

## Introduction

This handout accompanies Chapter 9 in [O'Sullivan and Unwin (2010)](http://www.wiley.com/WileyCDA/WileyTitle/productCd-0470288574.html). 

Here is how you can set up and use the continuous function on page 246. 

```{r}
z <- function(x, y) { -12 * x^3 + 10 * x^2 * y - 14 * x * y^2 + 25 * y^3 + 50 }
z(.5, 0.8)
```

Function `zf` adds some complexity to make it usable in the 'interpolate' function below.
```{r}
zf <- function(model, xy) {
   x <- xy[,1]
   y <- xy[,2]
   z(x, y)
}
```

Now use it

```{r fields5}
library(terra)
r <- rast(xmin=0.5, xmax=1.4, ymin=0.6, ymax=1.5, ncol=9, nrow=9, crs="")
z <- interpolate(r, model=NULL, fun=zf)
names(z) <- 'z'

vt <- persp(z, theta=30, phi=30, ticktype='detailed', expand=.8)
```

Note that persp returned something _invisibly_ (it won't be printed when not captured as a  variable, `vt`, in this case), the 3D transformation matrix that we use later. This is not uncommon in `R`. For example `hist` and `barplot` have similar behaviour.

```{r fields6}
pts <- as.data.frame(z, xy=TRUE)
pt <- trans3d(pts[,1], pts[,2], pts[,3], vt)
plot(pt, col=rainbow(9, .75, start=.2)[round(pts[,3]/10)-2], pch=20, cex=2)
```


For a more interactive experience, try:
```
library(rasterVis)
library(rgl)
# this opens a new window
plot3D(raster::raster(z), zfac=5)
```


We will be working with temperature data for California. You can [download](/data/precipitation.csv) the climate data used in the examples. 

```{r fields10}
library(rspat)
d <- spat_data("precipitation")
head(d)
d$prec <- rowSums(d[, c(6:17)])
plot(sort(d$prec), ylab='Annual precipitation (mm)', las=1, xlab='Stations')
```


```{r fields15}
dsp <- vect(d, c("LONG", "LAT"), crs= "+proj=longlat +datum=WGS84")
CA <- spat_data("counties")

cuts <- c(0,200,300,500,1000,3000)
blues <- colorRampPalette(c('yellow', 'orange', 'blue', 'dark blue'))
plot(CA, col="light gray", lwd=4, border="white")
plot(dsp, "prec", breaks=cuts, col=blues(5), pch=20, cex=1.5, add=TRUE, plg=list(x="topright"))
lines(CA, lwd=1.5)
```


