# Parallelization

## Introduction

Many computations on `SpatRaster` (and to a lesser extent `SpatVector`)  objects are slow because of the large (file) size of the raster. The computations are also embarrassingly parallel: the same operation is applied to many cells, many features, or many tiles, and the work for each piece is independent. There are several ways to take advantage of multiple cores or multiple machines when working with terra, and the right one depends on (a) the nature of the operation, (b) whether you are on a single computer or on a cluster, and (c) whether the data are in memory, in a single file, or scattered across many files.

In some cases parallelization is essential and easy --- for example when processing many files on a cluster, or for tasks that are repeated often enough that the setup cost is amortised. In other cases the gain is marginal and not worth the effort, because setting up a parallel workflow can take a long time, introduce subtle errors, and complicate debugging. It pays to first time a single-threaded run, estimate the best-case speed-up, and only then decide whether to parallelize. 

This chapter discusses the available approaches and their trade-offs, and ends with a roadmap for further parallelization support in terra.


## A fundamental constraint: terra objects hold a C++ pointer

`SpatRaster`, `SpatVector`, `SpatRasterDataset`, `SpatRasterCollection`, `SpatExtent` and `SpatGraticule` are *thin* R objects: each holds an external pointer to an underlying C++ object that lives in the memory of the R session that created it. As a consequence, these objects
**cannot be serialized** or sent to a worker process with `parallel::clusterExport` / `parLapply` / `mclapply` / `future` directly. After serialization the pointer would be a dangling address in the worker's memory.

If you try, the worker either errors out, returns `NULL`, or crashes:

```{r, eval=FALSE}
library(terra)
library(parallel)
r <- rast(system.file("ex/elev.tif", package="terra"))
cl <- makeCluster(2)
parLapply(cl, 1:2, function(i) global(r, "mean", na.rm=TRUE))   # FAILS
```

There are four ways out, described in the next sections.


## Strategy 1 — `wrap()` and `unwrap()`

`wrap()` converts a terra object into a regular R object (called `PackedSpatRaster`, `PackedSpatVector`, etc.) that holds the data and the
metadata in plain R memory. Such packed objects can be sent to workers, or returned across `future` boundaries. On the worker side, `unwrap()` reconstructs a usable `SpatRaster` / `SpatVector`. The terra version of `saveRDS` writes such packed objects to disk.

```{r, eval=FALSE}
library(terra)
library(parallel)

r <- rast(system.file("ex/elev.tif", package="terra"))
pr <- wrap(r)                      # serializable

cl <- makeCluster(4)
clusterEvalQ(cl, library(terra))   # workers need the package loaded

results <- parLapply(cl, 1:4, function(i, p) {
    x <- unwrap(p)                 # rebuild the SpatRaster on the worker
    global(x, "mean", na.rm=TRUE)
}, p = pr)

stopCluster(cl)
```

`wrap()` reads all values into memory, so it is most useful for small to medium rasters. For large rasters consider Strategy 2 instead.


## Strategy 2 — pass filenames, open on the worker

When the data already live in a file (GeoTIFF, NetCDF, ...) the cheapest "serialization" is the filename. Each worker opens its own `SpatRaster` from the file. terra is fully thread-safe at the file-open level (each worker gets its own dataset handle), and only the grid cells that are actually read are pulled from disk.

```{r, eval=FALSE}
fname <- system.file("ex/elev.tif", package="terra")
cl <- makeCluster(4)
clusterEvalQ(cl, library(terra))

results <- parLapply(cl, 1:4, function(i, f) {
    r <- rast(f)                   # cheap: only metadata is loaded
    global(r, "mean", na.rm=TRUE)
}, f = fname)

stopCluster(cl)
```

This pattern is clearly better than `wrap()` whenever:

- the dataset is too large to comfortably fit in worker memory,
- you have many files (one per worker, or one per task), or
- you are on an HPC cluster where all nodes can see the same shared file system.


## Strategy 3 — partition by tiles with `window()`

`window(x) <- ext(...)` sets a *read window*: subsequent operations on `x` behave as if `x` had only that extent, but no data are copied or moved. This is the natural way to split a large raster into tiles that are processed independently and then mosaicked back together.

```{r, eval=FALSE}
fname <- system.file("ex/elev.tif", package="terra")
r <- rast(fname)

tiles <- getTileExtents(r, 50, extend=FALSE, buffer=0)

cl <- makeCluster(4)
clusterEvalQ(cl, library(terra))

outs <- parLapply(cl, tiles, function(e, f) {
    x <- rast(f)
    window(x) <- e
    out <- focal(x, w=11, fun="mean")
    wrap(out) # return the small result
}, f = fname)

stopCluster(cl)

result <- do.call(mosaic, lapply(outs, unwrap))
```
In stead of returning the output, you can also write the results to files and combine these with `vrt` 

When tiles overlap (as they should for any window-based focal/aggregate operation, by at least the half-window size) the boundary pixels can later be discarded. 


## Strategy 4 — exchange raw values

For some workflows it is simplest to step out of the terra type system entirely: extract the cell values as a plain R object (a `numeric` vector, `matrix`, or `array`), send those numbers to the workers, do the computation, ship the numbers back, and stitch them into a `SpatRaster` on the master. Plain numeric data has no C++ pointer attached, so it serializes through any parallel back-end (`parLapply`, `mclapply`, `future`, `mirai`, `Rmpi`, ...) without `wrap()`.

For rasters that fit comfortably in memory you can extract everything at once with `values()` (returns a matrix with one column per layer) or `as.array()` (returns a 3-D array `[row, col, layer]`), split the data along whichever dimension makes sense, process the pieces in parallel, and assign the result back with `values(x) <- ...`.

```{r, eval=FALSE}
r <- rast(system.file("ex/elev.tif", package="terra"))
v <- values(r)                                  # plain matrix
chunks <- split(seq_len(nrow(v)), cut(seq_len(nrow(v)), 4))

cl <- makeCluster(4)
parts <- parLapply(cl, chunks, function(i, v) {
    sqrt(v[i, , drop=FALSE])                    # any vectorized R code
}, v = v)
stopCluster(cl)

out <- rast(r)
values(out) <- do.call(rbind, parts)
```

For rasters that do *not* fit in memory the same idea applies, but you process one block at a time using terra's chunking machinery (`readStart` / `readValues` / `writeStart` / `writeValues` / `writeStop`). Each block read returns a numeric vector; you can dispatch that vector to a worker, receive the result, and write it straight to the output file. The workers never see a `SpatRaster`, only numbers.

```{r, eval=FALSE}
r   <- rast("/data/big.tif")
out <- rast(r)
b   <- writeStart(out, "/data/big_out.tif", overwrite=TRUE)

readStart(r); on.exit(readStop(r), add=TRUE)
cl <- makeCluster(4)
on.exit(stopCluster(cl), add=TRUE)

# read all blocks, dispatch them in parallel, write the results in order
blocks  <- lapply(seq_len(b$n), function(i) readValues(r, b$row[i], b$nrows[i]))
results <- parLapply(cl, blocks, function(v) sqrt(v))
for (i in seq_len(b$n)) writeValues(out, results[[i]], b$row[i], b$nrows[i])
writeStop(out)
```

This approach is attractive when:

- the per-pixel computation is a fast vectorized R expression (or a call into a third-party numerical library that already knows how to do BLAS/OpenMP threading),
- you want to use a non-terra parallel back-end (`future`, `mirai`, `BiocParallel`, ...) without adding a `wrap()`/`unwrap()` step,
- the workers should not need terra installed at all.

For very large rasters you can combine it with Strategy 3 (read the values from a windowed `SpatRaster`) to keep memory bounded while still avoiding any `wrap()` round-trip.


## Within-process parallelism: TBB

A growing number of terra's C++ code paths are parallelized internally with [Intel TBB](https://github.com/oneapi-src/oneTBB). When TBB is
available at build time (which is the default on Windows and on most Linux distributions) you can opt into TBB-based threading by setting the `parallel` write option (this is the default in the version >= 1.9-21):

```{r, eval=FALSE}
# is TBB available?
terra:::.have_TBB()                 

# per call
focal(r, w=21, fun="mean", wopt=list(parallel=TRUE))

# session-wide default
terraOptions(parallel=TRUE)
focal(r, w=21, fun="mean")          
```

By default TBB schedules across all logical CPU cores. To cap the number of threads (for example to leave room for other processes, or to keep a forked worker from spawning hundreds of children), set the `threads` option. `threads = 0` (the default) means "no cap, use the TBB default". Any positive value caps the number of threads at that count. The option is honoured by every TBB-parallel kernel in terra and is also forwarded to GDAL's warp engine (`project()`, `resample()`, `warp()`).

```{r, eval=FALSE}
terraOptions(parallel=TRUE, threads=4)         # use at most 4 threads
focal(r, w=21, fun="mean")                     # capped at 4

terraOptions(threads=0)                        # back to "all CPUs"
```

Functions that currently benefit from TBB include parts of `arith`, distance calculations on `SpatVector`, and the new fast `focal` path
(`new=TRUE`, the default). TBB has no per-call setup cost and shares the process memory, so it is the most efficient way to use multiple cores **on a single machine** for one operation. Combine with Strategy 2 or 3 to reach across machines.


## Function-level parallelism: the `cores` argument

Some functions that apply a user-supplied R function over cells or layers take a `cores` argument and use `parallel::makeCluster` (PSOCK) under the hood. Currently this includes:

- `app(x, fun, ..., cores)`
- `tapp(x, index, fun, ..., cores)`
- `lapp(x, fun, ..., cores)`
- `predict(model, ..., cores)`  (and `interpolate`, `regress`)
- `aggregate(x, fact, fun, ..., cores)`

```{r, eval=FALSE}
r <- rast(system.file("ex/elev.tif", package="terra"))
out <- app(r, fun=function(v) median(v, na.rm=TRUE), cores=4)
```

terra creates the cluster, ships the function (and any extra named arguments) to each worker, and processes the raster block-by-block with
the workers in parallel. For `predict` you can also pass `cpkgs=` to preload packages on each worker, which is necessary when the prediction function comes from a package such as **randomForest** or **mgcv**.

Important caveats:

- Extra `...` arguments **must be named** when `cores > 1`, because the workers receive them by name through `clusterExport`.
- The function `fun` is serialized to each worker; it should not capture large objects from its enclosing environment, and any terra objects it uses must be passed in via `wrap()` or as filenames.
- The cluster is created and torn down per call. For repeated calls with many small tasks create your own cluster and pass that to argument `cores`

## Passing user functions: `app`, `lapp`, `tapp`, `focal`, `focalCpp`

Several terra functions accept a user-defined R function:

- `app(x, fun, ...)` applies `fun` to all values of each cell across
  layers.
- `lapp(x, fun, ...)` applies `fun` to layers by name.
- `tapp(x, index, fun, ...)` aggregates layers by group.
- `focal(x, w, fun, ...)` applies `fun` to values in a moving window.
- `focalCpp(x, w, fun, ...)` does the same but expects a vectorized
  function written for performance (e.g. via Rcpp).

For built-in functions (`"sum"`, `"mean"`, `"min"`, `"max"`, `"modal"`, "first"`) terra dispatches to the optimized C++ implementation and
parallelism is controlled via TBB (`wopt=list(parallel=TRUE)`). For arbitrary R functions the work happens in R, and the `cores` argument is the relevant knob (see previous section).


## Single computer vs. HPC cluster

The right choice depends on where the cores are.

### One computer, multiple cores

In rough order of preference:

1. **TBB** (`wopt=list(parallel=TRUE)`) — zero setup cost, no serialization, lowest overhead. Use whenever the operation supports it.
2. **`cores=` argument** in `app`/`lapp`/`tapp`/`predict`/etc. — for user-defined R functions on a single raster.
3. **Manual `parallel::parLapply` with `wrap()` or filenames** — for custom workflows that loop over rasters, layers, models, or tiles.
4. **`window()` tiling + `parLapply`** — when one operation on one large raster can be split spatially. Particularly effective for `focal`, `aggregate`, and `predict` on rasters that do not fit in memory.
5. **Raw values via `values()` / `readValues`** — when the per-pixel computation is a fast vectorized R expression and you would rather not move terra objects across process boundaries at all.

`mclapply` and `future::multicore` (forked workers) work too on macOS and Linux. They share memory copy-on-write so `wrap()` is unnecessary, but the C++ pointer must still be valid on the worker, which it is for forked processes that did not modify the object after the fork. Avoid them on Windows and in RStudio (they fall back to sequential).

### HPC cluster of nodes (each with multiple cores)

On an HPC system you typically combine **two levels** of parallelism:

- **Across nodes**: one R process per node (or per task), launched via the scheduler (SLURM, PBS, ...) using `Rscript`, `batchtools`,
  `future.batchtools`, MPI (via `Rmpi`/`pbdMPI`), or simply by partitioning input filenames or tiles across array jobs.
- **Within node**: TBB (set `terraOptions(parallel=TRUE)` at the top of the worker script) and/or `cores=` for user functions.

The most reliable inter-node "serialization" of raster data is **the filename on a shared file system** (Lustre, NFS, GPFS, ...). Build the
SpatRaster on each node from the path; never `wrap()` and ship raster values across the network if the file is reachable.

```{r, eval=FALSE}
# slurm_array.R, dispatched as one task per tile
args <- commandArgs(trailingOnly=TRUE)
i <- as.integer(args[1])

library(terra)
terraOptions(parallel=TRUE)         # use TBB within this node

r <- rast("/shared/dem.tif")
window(r) <- tiles[[i]]             # tiles defined in a shared script
out <- focal(r, w=11, fun="mean")
writeRaster(out, sprintf("/shared/out/tile_%03d.tif", i))
```

Then mosaic the tiles in a final, single-process step.


## Other useful patterns

- **`tiles()`** generates `SpatExtent` tiles from a template raster — a convenient input for the tiling strategy above.
- **`makeTiles(x, y)`** writes one file per tile to disk; a cluster job can then process the tiles independently.
- **`vrt()`** stitches many files back into a single virtual raster without copying any pixels.
- **`writeRaster(..., gdal=c("BIGTIFF=YES", "NUM_THREADS=ALL_CPUS"))`** uses GDAL's own threading for compression.
- **`spatSample()`** with a large `size` runs faster with TBB.
- For **`SpatVector`**: most geometric predicates (`relate`, `nearest`, `distance`) honour TBB; for very large layers consider partitioning
  the layer geographically with `crop()` and processing the parts on separate workers.


## Roadmap for better parallelization support

Ideas for areas where terra can be improved:

1. **More TBB coverage**.
   
   Obvious next targets are `aggregate(fun=...)`, `extract`, `terrain`, `mask` / `cover`, `classify`, `subst`, `crop`, `extend`, `predict` (built-in stat models), and the `merge`/`mosaic` core. Many of these are pixel-wise or block-wise operations whose inner loops are immediately parallelizable.

2. **Block-level parallelism in writeRaster.**
   
   Most chunked writers process blocks sequentially. With a thread-safe block queue and an output GDAL dataset opened in `UPDATE` mode,
   blocks could be computed and written in parallel.

3. **A `tile_apply()` convenience wrapper.**
   
   A single high-level function that takes `(x, fun, tiles, cores, cpkgs, ...)`, runs `fun` on each tile (with `window()` set under the
   hood) on a worker cluster, and returns a mosaicked `SpatRaster`. This would package the Strategy 3 pattern shown above.

4. **First-class `future`/`mirai` integration.**
   
   `wrap()` plus `future_lapply()` already works, but a thin `terra::future_app()` / `terra::future_focal()` would handle the
   wrap/unwrap dance and enforce filename-based shipping for large rasters automatically.

5. **GPU offload.**

   Several inner kernels (`focal`, `aggregate`, `arith`, `terrain`) are regular enough to map cleanly to CUDA / SYCL / OpenMP target offload. This is a larger undertaking but would dwarf the speedups achievable with multi-core CPUs for pixel-wise operations.

