---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Spatial prediction

Some examples for prediction. 

## Predict 


### GLM

```{r glm, message=FALSE}
library(terra)
logo <- rast(system.file("ex/logo.tif", package="terra"))   
names(logo) <- c("red", "green", "blue")
p <- matrix(c(48, 48, 48, 53, 50, 46, 54, 70, 84, 85, 74, 84, 95, 85, 
   66, 42, 26, 4, 19, 17, 7, 14, 26, 29, 39, 45, 51, 56, 46, 38, 31, 
   22, 34, 60, 70, 73, 63, 46, 43, 28), ncol=2)

a <- matrix(c(22, 33, 64, 85, 92, 94, 59, 27, 30, 64, 60, 33, 31, 9,
   99, 67, 15, 5, 4, 30, 8, 37, 42, 27, 19, 69, 60, 73, 3, 5, 21,
   37, 52, 70, 74, 9, 13, 4, 17, 47), ncol=2)

xy <- rbind(cbind(1, p), cbind(0, a))

# extract predictor values for points
e <- extract(logo, xy[,2:3])

# combine with response 
v <- data.frame(cbind(pa=xy[,1], e[,-1]))

#build a model, here with glm 
model <- glm(formula=pa~., data=v)

#predict to a raster
r1 <- predict(logo, model)

plot(r1)
points(p, bg='blue', pch=21)
points(a, bg='red', pch=21)

# logistic regression
model <- glm(formula=pa~., data=v, family="binomial")
r1log <- predict(logo, model, type="response")

# use a modified function to get the probability and standard error
# from the glm model. The values returned by "predict" are in a list,
# and this list needs to be transformed to a matrix

predfun <- function(model, data) {
  v <- predict(model, data, se.fit=TRUE)
  cbind(p=as.vector(v$fit), se=as.vector(v$se.fit))
}

r2 <- predict(logo, model, fun=predfun)
```


### Principal components

Here using sampling to simulate an object too large to feed all its values to prcomp

```{r pca}
sr <- values(spatSample(logo, 100, as.raster=TRUE))
pca <- prcomp(sr)

x <- predict(logo, pca)
plot(x)

library(pls)
model <- plsr(formula=pa~., data=v)
# this returns an array:
predict(model, v[1:5,])
# write a function to turn that into a matrix
pfun <- function(x, data) {
   y <- predict(x, data)
   d <- dim(y)
   dim(y) <- c(prod(d[1:2]), d[3])
   y
}

pp <- predict(logo, model, fun=pfun)
```


### Random Forest

```{r rf}
library(randomForest)
rfmod <- randomForest(pa ~., data=v)

## note the additional argument "type='response'" that is 
## passed to predict.randomForest
r3 <- predict(logo, rfmod, type='response')

## get class membership probabilities
vv <- v
vv$pa <- as.factor(vv$pa)
rfmod2 <- randomForest(pa ~., data=vv)
r4 <- predict(logo, rfmod2, type='prob')
rr4 <- as(r4, "Raster")
raster::spplot(rr4)
```

## cforest 

cforest is an alternative Random Forest implementation. Here an example with a `factors` argument

```{r cfor}
library(party)
m <- cforest(pa~., control=cforest_unbiased(mtry=3), data=v)
# the second argument in party:::predict.RandomForest
# is "OOB", and not "newdata" or similar. We need to write a wrapper
# predict function to deal with this 	
predfun <- function(m, d, ...) predict(m, newdata=d, ...)

pc <- predict(logo, m, OOB=TRUE, fun=predfun)

# knn, using "app" instead of "predict"
library(class)
cl <- factor(c(rep(1, nrow(p)), rep(0, nrow(a))))
train <- extract(logo, rbind(p, a))[, -1]
k <- app(logo, function(x) as.integer(as.character(knn(train, x, cl))))
```


## Interpolate


### Thin plate spline interpolation with x and y only

```{r tps, message=FALSE}
library(terra)
# example data
r <-rast(system.file("ex/test.tif", package="terra"))
ra <- aggregate(r, 10)
xy <- data.frame(xyFromCell(ra, 1:ncell(ra)))
v <- values(ra)

# Thin plate spline model
library(fields) 
tps <- Tps(xy, v)
x <- rast(r)

# use model to predict values at all locations
p <- interpolate(x, tps)
p <- mask(p, r)
plot(p)

# change the fun from predict to fields::predictSE to get the TPS standard error
se <- interpolate(x, tps, fun=predictSE)
se <- mask(se, r)
plot(se)
```


Add another predictor variable; let"s call it elevation

```{r tps2, message=FALSE}
elevation <- (init(r, "x") * init(r, "y")) / 100000000
names(elevation) <- "elev"
elevation <- mask(elevation, r)

z <- extract(elevation, vect(xy), fun=function(x)x[1])
z <- z[,2,drop=FALSE]

# add as another independent variable
vv <- na.omit(cbind(xy, z, v))
tps2 <- Tps(vv[,1:3], vv[,4])
#p2 <- interpolate(elevation, tps2)
#plot(p2) 

# as a linear coveriate
tps3 <- Tps(vv[,1:2], vv[,4], Z=vv[,3])

# Z is a separate argument in Krig.predict, so we need a new function
# Internally (in interpolate) a matrix is formed of x, y, and elev (Z)

pfun <- function(model, x, ...) {
   predict(model, x[,1:2], Z=x[,3], ...)
}
p3 <- interpolate(elevation, tps3, fun=pfun)
plot(p3)
```

### kriging 

Kriging with gstat examples. Examples provided by Maurizio Marchi


```{r gstat}
library(sp)
library(gstat)
data(meuse)

# inverse distance weighted (IDW)
r <-rast(system.file("ex/test.tif", package="terra"))
mg <- gstat(id = "zinc", formula = zinc~1, locations = ~x+y, data=meuse, 
            nmax=7, set=list(idp = .5))
f <- function(model, data, ...) predict(model, data, ...)[,3,drop=FALSE]
z <- interpolate(r, mg, fun=f, debug.level=0)

# kriging
coordinates(meuse) <- ~x+y
crs(meuse) <- crs(r)

# ordinary kriging
v <- variogram(log(zinc)~1, meuse)
m <- fit.variogram(v, vgm(1, "Sph", 300, 1))
gOK <- gstat(NULL, "log.zinc", log(zinc)~1, meuse, model=m)

fg <- function(model, d, crs, ...) {
	sp <- SpatialPointsDataFrame(d[,1:2,drop=FALSE], data.frame(d), proj4string=CRS(crs))
	data.frame(predict(model, sp, ...))[,3:4]
}
OK1 <- interpolate(r, gOK, fun=fg, debug.level=0, crs=crs(r), na.rm=TRUE)
OK2 <- interpolate(r, gOK, fun=fg, debug.level=0, crs=crs(r), na.rm=FALSE)
OK3 <- interpolate(x, gOK, fun=fg, debug.level=0, crs=crs(x))
plot(c(OK1[[1]], OK2[[1]], OK3[[1]]))


## universial kriging
vu <- variogram(log(zinc)~elev, meuse)
mu <- fit.variogram(vu, vgm(1, "Sph", 300, 1))
gUK <- gstat(NULL, "log.zinc", log(zinc)~elev, meuse, model=mu)
names(r) <- "elev"
UK <- interpolate(r, gUK, fun=fg, debug.level=0, crs=crs(r))

## co-kriging
gCoK <- gstat(NULL, "log.zinc", log(zinc)~1, meuse)
gCoK <- gstat(gCoK, "elev", elev~1, meuse)
gCoK <- gstat(gCoK, "cadmium", cadmium~1, meuse)
gCoK <- gstat(gCoK, "copper", copper~1, meuse)
coV <- variogram(gCoK)
plot(coV, type="b", main="Co-variogram")
coV.fit <- fit.lmc(coV, gCoK, vgm(model="Sph", range=1000))
coV.fit
plot(coV, coV.fit, main="Fitted Co-variogram")
coK <- interpolate(r, coV.fit, fun=fg, debug.level=0, crs=crs(r))
plot(coK)
```

