# Downloading MODIS data

MODIS data can be acquired through multiple services, and it can be confusing to find out what to get where. This [website](https://modis.gsfc.nasa.gov/tools/) has a lot of information on MODIS resources. 

Websites are great for preliminary data exploration, they are not ideal for accessing large number of files or when automatic updates are required. Manually downloading also makes workflows less reproducible. So once you know what data you need, it is better to use automatic downloading methods. Here we show how to use *R* package `luna` to download MODIS data for a geographic area and time period. 

A large number of MODIS products is available. We first need to find data for the product that we deem best suited for our study.

```{r product, message=FALSE}
library(terra)
library(luna)

# lists all products that are currently searchable
prod <- getProducts() 
head(prod)

# to find the MODIS products
modis <- getProducts("^MOD|^MYD|^MCD")
head(modis)
```

We will use "MOD09A1" for this turorial.

```{r prod}
product <- "MOD09A1"
```

To learn more about a specific product you can launch a webpage

```
productInfo(product)
```

Note that the entire MODIS archive is regularly re-processed for overall imporvement and revisions. We use version 6 or later for our analysis.

Once we finalize the product we want to use, we define some parameters for the data we want: product name, start and end date, and area of interest.

```{r dates}
start <- "2010-01-01" 
end <- "2010-01-07"
```

We will download an example MODIS 8-day composite tile. Our area of interest is Marsabit county, Kenya. To define the area of interest, we can define a spatial extent, or use an object that has an extent. Here we use a polygon for Marsabit.


```{r aoi1}
ken <- geodata::gadm("Kenya", level=1, path=".") 
ken
```

`ken` is a `SpatVector` of polygons. We can subset it get the polygon for Marsabit:

```{r aoi2}
i <- ken$NAME_1 == "Marsabit"
aoi <- ken[i,]
```

And the check our results we make a map

```{r aoi3}
plot(ken, col="light gray")
lines(aoi, col="red", lwd=2)
```

Let's now find out what MODIS data is available for this area. We can search the data available from a NASA server

```{r getfilenames}
mf <- luna::getNASA(product, start, end, aoi=aoi, download = FALSE)
mf
```

To download the tiles, usually you would download them to a folder where you save the data for your project. Here we use the temporary directory. You should use a specific direcory of your choice instead.

```{r downloaddir}
datadir <- file.path(dirname(tempdir()), "_modis")
dir.create(datadir, showWarnings=FALSE)
```

You also need to provide the username and password for your (free) EOSDIS account. If you do not have an account, you can sign up [here](https://urs.earthdata.nasa.gov/users/new). My passwords are stored in a file that I read below (sorry, I cannot show you the values).

```{r pwd}
up <- readRDS("../../../../pwds.rds")
up <- up[up$service == "EOSDIS", ]
```

Now we are ready to download the data

```{r download}
mf <- luna::getNASA(product, start, end, aoi=aoi, download=TRUE,
                     path=datadir, username=up$user, password=up$pwd)
mf
```

In the next chapters we will use the downloaded files. 
	