```{r, echo=FALSE}
```
# Image exploration

## Introduction

Now that we have successfully downloaded one MODIS tile, we can use `terra` package to explore and visualize it. Please note that MODIS tiles are distributed in HDF format that may include sub-datasets. The sub-dataset and processing steps might be different for various MODIS collections (e.g. daily scenes, vegetation indices).

Now that we have donwload some MODIS data, we can explore and visualize it.

First create a SpatRaster object from the file created on the previous page.

```{r mplot, message=FALSE}
datadir <- file.path(dirname(tempdir()), "_modis")
mf <- file.path(datadir, "MOD09A1.A2009361.h21v08.006.2015198070255.hdf")

library(terra)
r <- rast(mf[1])
r
```

*Exercise:* Find out at least 5 properties (path, row, date of collection etc) of the MODIS data from the  information embedded in the filename.


## Image properties

The code below illustrates how you can load HDF files and access image properties of a SpatRaster object. 

The coordinate reference system (CRS)
```{r imgp1}
crs(r)
```

And the number of cells, rows, columns
```{r imgp2}
dim(r)
nrow(r)
ncol(r)
# Number of layers (bands)
nlyr(r)
ncell(r)
```

The spatial resolution is about 500 m
```{r imgp3}
res(r)
```

The layernames tell us what "bands" we have

```{r imgp5}
names(r)
```

## Plot

Now let's make some simple plots to see if things look reasonable. 

```{r plotrgb}

# Create an image RGB composite plot
plotRGB(r, r = 1, g = 4, b = 3)

# Disappointing? apply some stretching; see `?plotRGB` for more options

plotRGB(r, r = 1, g = 4, b = 3, stretch="lin")
```

*Exercise:* Create False Color Composite plot using the same data. Hint: try `plotRGB` and specify the bands you need.    

*Exercise:* Save the plots to grahics files. Hint: have a look at `?png`.
