# Quality control

Not all pixels in a MODIS image are not suitable for analysis related to land surafces. For example, pixels can be affected by cloud or other atmospheric conditions. For that reason, each image needs to be pre-processed to remove non-desirable pixels. 

Each MODIS tile contains quality assurance (QA) information that can be used to identify the pixels affected by, for example, clouds, and shadows. However the QA information are stored in bit-encoding format to represent different possibile combinations of quality issues. In binary notation, a single bit represent two values --- 0 (no) or 1(yes), two bits represent four values (all possible combination of 0 & 1: 00, 01, 10, 11). [This video by USGS](https://www.usgs.gov/media/videos/getting-started-modis-v6-surface-reflectance-data-part-3) explains the QA encoding.     

The table below shows State QA description (16-bit) for 500 m, 1 km and coarser resolution MODIS surface reflectance products. For product specific QA attribute, follow the [MODIS Surface Reflectance User’s Guide](https://lpdaac.usgs.gov/documents/306/MOD09_User_Guide_V6.pdf).

```{r StateQA, echo=FALSE}
table <- read.csv("sur_refl_state.txt", header=TRUE, sep = "\t", colClasses = "character", encoding ="UTF-8")
dm <- data.frame(table, stringsAsFactors = FALSE)
colnames(dm) <- c("bit", "variable", "value", "description")
dm[,4] = gsub("\n", " ", dm[,4])
knitr::kable(dm, align="l")
```

Note that in R, the first bit will be referred as "1", whereas in many other languages (e.g. Python) it will follow the values as shown in table.  

To interpret the pixel level QA values, we need to convert them from decimal to binary format. The `luna` package offers a function for this conversion and create a mask from the QA band. The user need to specify a matrix ("qabits") with the start and end of the quality assessment (QA) bits considered, and specify a list ("reject") with the values to be rejected (in the image) matching the rows in `qabits`. Following the table above, we will define the "reject" values to exclude pixels affected by cloud and cloud shadow.  

```{r qcconst}
from <- c(1,3,11,14)
to   <- c(2,3,11,14)
reject <- c("01,10", "1", "1", "1")

qa_bits <- cbind(from, to, reject)
qa_bits
```

Pixels with bits 1 and 2 with values "01" or "10" will be rejected. All other combinations ("00" and "11" in this case) are not rejected. 

We use the downloaded MODIS file that, in a previouse step, we saved in the `datadir` directory.

```{r mod, message=FALSE}
library(terra)
datadir <- file.path(dirname(tempdir()), "modis")
mf <- file.path(datadir, "MOD09A1.A2009361.h21v08.006.2015198070255.hdf")

r <- rast(mf)
```

Generate the quality mask. We will use band 12 `sur_refl_state_500m` that has the quality data.

```{r qc1}
qc <- r[[12]]

plot(qc, main = "Quality")
```

The luna package has a `modis_mask` method to create a mask from the quality band and the parameters defined above. 

```{r qc2}
library(luna)

quality_mask <- modis_mask(qc, 16, qa_bits)

plot(quality_mask, main="Quality mask")  
```

The plot shows the pixels we want to retain. Now that we have the quality mask, we can apply it to all the bands. It is always useful to visually check the result of the masking as many times pixels could be wrongly flagged as "poor" quality. Residual noises can be filtered in subsequent gap-filling and smoothing operations.   

```{r qcmask}
rmask <- mask(r, quality_mask)
```

And we can plot the results, here as a "false color composite" (NIR:Red:Green)

```{r qcmaskplot}
plotRGB(rmask, r = 2, g = 1, b = 4, main='False color composite', stretch="lin")
```

Finally we save the result after cloud masking.

```{r savefile, echo=FALSE}
mf <- file.path(datadir, "modis_qualmasked.tif")

writeRaster(rmask, mf, overwrite=TRUE)
```

