
```{r, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60))
datadir <- file.path(dirname(tempdir()), "agrin/modis")
mf <- file.path(datadir, "MOD09A1.A2009361.h21v08.006.2015198070255.hdf")
library(terra)
```

# Quality control

We will now mask out bad quality pixels (e.g. clouds and shadows)

Not all pixels in a MODIS image are of good quality. We need to pre-process the data to remove bad pixels. 

To do so we can create a mask for e.g., clouds, shadows, and pixels with low quality, using Surface reflectance 500 m state quality assessment (QA) bits (see Table below).

500 m, 1 km and Coarse Resolution Surface Reflectance Band Quality Description (32-bit). 

```{r StateQA, echo=FALSE}
table <- read.table("MODIS_State_QA.txt", header=TRUE, sep = "\t", colClasses = "character", encoding ="UTF-8")
dm <- data.frame(table, stringsAsFactors = FALSE)
colnames(dm) <- c("bit", "variable", "value", "description")
dm[,4] = gsub("\n", " ", dm[,4])
knitr::kable(dm, align="l")
```

Specify a matrix ("se") with the start and end of the quality assessment (QA) bits considered, and specify a list ("reject") with the values to be rejected (in the image) matching the rows in `se`.

```{r qcconst}
se <- matrix(c(1,2,3,6,11,11,12,14,16,16), ncol=2, byrow=TRUE)
reject <- list(c("10", "11"), c("1100","1101","1110","1111"), "1", c("000","110","111"), "11")
se
str(reject)
```

Thus, if bytes 1 and 2 have values "10" or "11", these are rejected. All other combinations ("00" and "01" in this case) are not rejected.


We use the downloaded MODIS file that we save in the `datadir` directory.

```{r mod}
library(terra)
mf 
r <- rast(mf)
```

Generate the quality mask. We will use band 12 `sur_refl_state_500m` that has the quality data.

```{r qc1}
qc <- r[[12]]

plot(qc, main = "Quality")
```

The luna package has the `modis_mask` method creates a mask from the quality band and the parameters defined above. 

```{r qc2}
library(luna)

quality_mask <- modis_mask(qc, 16, se, reject)

plot(quality_mask, main="Quality mask with pixles to be retained")  
```

Now that we have the quality mask, we cna apply it to all the bands. 

```{r qcmask}
rmask <- mask(r, quality_mask)
```

And we can plot the results, here as a "false color composite" (NIR:Red:Green)

```{r qcmaskplot}
plotRGB(rmask, r = 2, g = 1, b = 4, main='False color composite', stretch="lin")
```

Finally we save the result after cloud masking.

```{r savefile, echo=FALSE}
datadir <- file.path(dirname(tempdir()), "agrin/modis")

mf <- file.path(datadir, "modis_qualmasked.tif")

writeRaster(rmask, mf, overwrite=TRUE)
```

