# Processing

Here we crop an image to get an exact area of interest, and then we compute the Normalized Difference Vegetation Index (NDVI).

Get the polygon boundary

```{r boundary, messages=FALSE}
library(terra)
library(geodata)
ken <- gadm(country="KEN", level=1, path=".")
pol <- ken[ken$NAME_1 == "Marsabit", ]
```


Change vector boundary coordinate reference system, so that it matches that of the MODIS data. 

```{r prj}
datadir <- file.path(dirname(tempdir()), "_modis")
mf <- file.path(datadir, "modis_qualmasked.tif")
rmask <- rast(mf)

prj <- crs(rmask)

prj

poly <- project(pol, prj)
```

**Question**: Why do not we change the coordinate system of the MODIS data?


Crop the image using the transformed vector boundaries.

```{r crop}
rcrop <- crop(rmask, poly)
```

Plot cropped MODIS and add the boundary. We use plotRGB to make a *false color composite* (near-infrared, red, green)

```{r cropmap}
plotRGB(rcrop, r = 2, g = 1, b = 4,  main = "False color composite", stretch = "lin" )
lines(poly, col="blue")
```


## NDVI

We have so far masked out bad quality pixels and clipped the image to area of interest extents. Let us now use the processed image to compute an index measure. The Normalized Difference Vegetation Index (NDVI) is a common measure of greenness. It is computed as follows

$$
NDVI = \frac{(NIR - Red)}{(NIR + RED)}.
$$


We expect the reflectance to be between 0 (very dark areas) and 1 (very bright areas). Due to various reasons, there may values slightly outside this rnage. First clamp values of the image between 0 and 1. 

```{r clamp}
r <- clamp(rcrop, 0, 1) 
```

```{r ndvi}
ndvi <- (r[[2]] - r[[1]]) /(r[[2]] + r[[1]])
plot(ndvi, main="NDVI")
```

*Exercise* Write a function to compute `NDVI` type two-band spectral indices and compute `NDVI` using the function.




