# Vector data manipulation

This chapter illustrates some ways in which we can manipulate vector data. We start with an example SpatVector that we read from a shapefile.

```{r, vec1}
library(terra)
f <- system.file("ex/lux.shp", package="terra")
p <- vect(f)
p
```

We can plot these data in many ways. For example:

```{r}
plot(p, "NAME_2")
```

## Basics

### Geometry and attributes

To extract the attributes (data.frame) from a SpatVector, use:
```{r}
d <- as.data.frame(p)
head(d)
```

You can also extract the geometry as a a matrix (this is rarely needed).

```{r}
g <- geom(p)
head(g)
```

Or as "well-known-text" 

```{r}
g <- geom(p, wkt=TRUE)
substr(g, 1, 50)
```


### Variables

You can extract a variable as you would do with a `data.frame`.

```{r}
p$NAME_2
```

To sub-set a SpatVector to one or more variables you can use the notation below. Note how this is different from the above example. Above a vector of values is returned. With the approach below you get a new SpatVector with only one variable. 

```{r}
p[, "NAME_2"]
```

You can add a new variable to a `SpatVector` just as if it were a `data.frame`.

```{r}
set.seed(0)
p$lets <- sample(letters, nrow(p))
p
```

Note that to get the number of geometries of SpatVector `p`, you can use `nrow(p)`, or `size(p)`. You can also do `perim(p)` to get the "length" of the spatial objects (zero for points, the length of the lines, or the perimeter of the polygons).


```{r}
perim(p)
```

Assigning a new value to an existing variable.

```{r}
p$lets <- sample(LETTERS, nrow(p))
head(p)
```

To get rid of a variable, set it to `NULL`.

```{r}
p$lets <- NULL
```

### Merge

You can assign an attributes table (data.frame) to a SpatVector with `values<-`. To add attributes to a SpatVector that already has attributes use `merge` (or `cbind` if you know the order of the records is the same).

```{r}
dfr <- data.frame(District=p$NAME_1, Canton=p$NAME_2, Value=round(runif(length(p), 100, 1000)))

dfr <- dfr[order(dfr$Canton), ]
pm <- merge(p, dfr, by.x=c('NAME_1', 'NAME_2'), by.y=c('District', 'Canton'))
pm
head(pm)
```

Note the new variable `Value` added to `pm`


### Records

Selecting rows (records).

```{r recs}
i <- which(p$NAME_1 == 'Grevenmacher')
g <- p[i,]
g
```

It is also possible to interactively select and query records by clicking on a plotted dataset. That is difficult to show here. See `?sel` for interactively selecting geometries and `?click` to identify attributes by clicking on a plot (map).


## Append and aggregate


## Append

More example data. Object `z` consists of four polygons; `z2` is one of these four polygons.

```{r, zzz}
z <- rast(p)
dim(z) <- c(2,2)
values(z) <- 1:4
names(z) <- 'Zone'
# coerce SpatRaster to SpatVector polygons
z <- as.polygons(z)
z
z2 <- z[2,]
plot(p)
plot(z, add=TRUE, border='blue', lwd=5)
plot(z2, add=TRUE, border='red', lwd=2, col='red')
```


To append SpatVector objects of the same (vector) type you can use `rbind`

```{r}
b <- rbind(p, z)
head(b)
tail(b)
```

Note how `rbind` allows you to append SpatVector objects with different attribute names, unlike the standard `rbind` for `data.frame`s. 


## Aggregate

It is common to aggregate ("dissolve") polygons that have the same value for an attribute of interest. In this case, if we do not care about the second level subdivisions of Luxembourg, we could aggregate by the first level subdivisions.

```{r, agg}
pa <- aggregate(p, by='NAME_1')
za <- aggregate(z)
plot(za, col='light gray', border='light gray', lwd=5)
plot(pa, add=TRUE, col=rainbow(3), lwd=3, border='white')
```

It is also possible to aggregate polygons without dissolving the borders.

```{r, aggnodis}
zag <- aggregate(z, dissolve=FALSE)
zag
plot(zag, col="light gray")
```

This is a structure that is similar to what you may get for an archipelago: multiple polygons represented as one entity (one row). Use `disagg` to split these up into their parts.

```{r diis}
zd <- disagg(zag)
zd
```


## Overlay

There are many different ways to "overlay" vector data. Here are some examples:

### Erase

Erase a part of a SpatVector
```{r raser}
e <- erase(p, z2)
plot(e)
```

### Intersect

Intersect SpatVectors

```{r, int}
i <- intersect(p, z2)
plot(i)
```

You can also `intersect` or `crop` with a SpatExtent (rectangle). The difference between `intersect` and `crop` is that with `crop` the geometry of the second argument is not added to the output.

```{r, intext}
e <- ext(6, 6.4, 49.7, 50)
pe <- crop(p, e)
plot(p)
plot(e, add=TRUE, lwd=3, col="red")
plot(pe, col='light blue', add=TRUE)
plot(e, add=TRUE, lwd=3, border="blue")
```


### Union

Get the union of two SpatVectors.

```{r}
u <- union(p, z)
u
```

Note that there are many more polygons now. One for each unique combination of polygons (and attributes in this case).

```{r unionplot}
set.seed(5)
plot(u, col=sample(rainbow(length(u))))
```

### Cover

`cover` is a combination of `intersect` and `union`. `intersect` returns new (intersected) geometries with the attributes of both input datasets. `union` appends the geometries and attributes of the input. `cover` returns the intersection and appends the other geometries and attributes of both datasets.

```{r, cov}
cov <- cover(p, z[c(1,4),])
cov
plot(cov)
```


### Difference


The symmetrical difference of two SpatVectors

```{r, dif}
dif <- symdif(z,p)
plot(dif, col=rainbow(length(dif)))
dif
```

## Spatial queries

We can query polygons with points ("point-in-polygon query"). 

```{r, pts}
pts <- matrix(c(6, 6.1, 5.9, 5.7, 6.4, 50, 49.9, 49.8, 49.7, 49.5), ncol=2)
spts <- vect(pts, crs=crs(p))
plot(z, col='light blue', lwd=2)
points(spts, col='light gray', pch=20, cex=6)
text(spts, 1:nrow(pts), col='red', font=2, cex=1.5)
lines(p, col='blue', lwd=2)
```

`extract` is used for queries between SpatVector and SpatRaster objects, and also for queries between SpatVectors.

```{r}
extract(spts, p)
extract(spts, z)
```

