# Vector data manipulation

This chapter illustrates some ways in which we can manipulate vector data. We start with an example SpatVector that we read from a shapefile.

```{r, vec1}
library(terra)
f <- system.file("ex/lux.shp", package="terra")
p <- vect(f)
p
```

We can plot these data in many ways. For example:

```{r}
plot(p, "NAME_2")
```

## Basics

### Geometry and attributes

To extract the attributes (data.frame) from a SpatVector, use:
```{r}
d <- as.data.frame(p)
head(d)
```

You can also extracting the geometry (rarely needed) as a a matrix.

```{r}
g <- geom(p)
head(g)
```

Or as "well-known-text" 

```{r}
g <- geom(p, wkt=TRUE)
substr(g, 1, 50)
```


### Variables

You can extracting a variable as you would do with a `data.frame`.

```{r}
p$NAME_2
```

To sub-setting a SpatVector to one or more variables you can use the notation below. Note how this is different from the above example. Above a vector of values is returned. With the approach below you get a new SpatVector with only one variable. 

```{r}
p[, "NAME_2"]
```

You can add a new variable to a `SpatVector` just as if it were a `data.frame`.

```{r}
set.seed(0)
p$lets <- sample(letters, nrow(p))
p
```

Note that to get the number of geometries of SpatVector `p`, you can use `nrow(p)`, or `size(p)`. You can also do `perimeter(p)` to get the "length" of the spatial objects (zero for points, the length of the lines, or the perimeter of the polygons).


```{r}
perimeter(p)
```

Assigning a new value to an existing variable.

```{r}
p$lets <- sample(LETTERS, nrow(p))
head(p)
```

To get rid of a variable, set it to `NULL`.

```{r}
p$lets <- NULL
```

### Merge

You can assign an attributes table (data.frame) to a SpatVector with `values<-`. To add attributes to a SpatVector that already has attributes use `merge`.

```{r}
dfr <- data.frame(District=p$NAME_1, Canton=p$NAME_2, Value=round(runif(length(p), 100, 1000)))

dfr <- dfr[order(dfr$Canton), ]
pm <- merge(p, dfr, by.x=c('NAME_1', 'NAME_2'), by.y=c('District', 'Canton'))
pm
head(pm)
```

Note the new variable `Value` added to `pm`


### Records

Selecting rows (records).

```{r recs}
i <- which(p$NAME_1 == 'Grevenmacher')
g <- p[i,]
g
```

It is also possible to interactively select and query records by clicking on a plotted dataset. That is difficult to show here. See `?select` for interactively selecting geometries and `?click` to identify attributes by clicking on a plot (map).


## Append and aggregate


## Append

More example data. Object `z`, consisting of four polygons, and `z2` which is one of these four polygons.

```{r, zzz}
z <- rast(p)
dim(z) <- c(2,2)
values(z) <- 1:4
names(z) <- 'Zone'
# coerce SpatRaster to SpatVector polygons
z <- as.polygons(z)
z
z2 <- z[2,]
plot(p)
plot(z, add=TRUE, border='blue', lwd=5)
plot(z2, add=TRUE, border='red', lwd=2, col='red')
```


To append SpatVector objects of the same (vector) type you can use `c`

```{r}
b <- c(p, z)
head(b)
tail(b)
```

Note how `c` allows you to append `SpatVect` objects with different attribute names. 


## Aggregate

```{r, agg}
pa <- aggregate(p, by='NAME_1')
za <- aggregate(z)
plot(za, col='light gray', border='light gray', lwd=5)
plot(pa, add=TRUE, col=rainbow(3), lwd=3, border='white')
```


Aggregate without dissolve
```{r, aggnodis}
zag <- aggregate(z, dissolve=FALSE)
zag
plot(zag, col="light gray")
```

This is a structure that is similar to what you may get for an archipelago: multiple polygons represented as one entity (one row). Use `disaggregate` to split these up into their parts.

```{r diis}
zd <- disaggregate(zag)
zd
```


## Overlay

### Erase

Erase a part of a SpatVector
```{r raser}
e <- erase(p, z2)
plot(e)
```

### Intersect

Intersect SpatVectors

```{r, int}
i <- intersect(p, z2)
plot(i)
```

You can also intersect with a SpatExtent (rectangle). 

```{r, intext}
e <- ext(6, 6.4, 49.7, 50)
pe <- crop(p, e)
plot(p)
plot(e, add=TRUE, lwd=3, col="red")
plot(pe, col='light blue', add=TRUE)
plot(e, add=TRUE, lwd=3, border="blue")

```


### Union

Get the union of two SpatVectors.

```{r}
u <- union(p, z)
u
```

Note that there are many more polygons now. One for each unique combination of polygons (and attributes in this case).

```{r unionplot}
set.seed(5)
plot(u, col=sample(rainbow(length(u))))
```

### Cover

`cover` is a combination of `intersect` and `union`. `intersect` returns new (intersected) geometries with the attributes of both input datasets. `union` appends the geometries and attributes of the input. `cover` returns the intersection and appends the other geometries and attributes of both datasets.

```{r, cov}
cov <- cover(p, z[c(1,4),])
cov
plot(cov)
```


### Difference


The symmetrical difference of two SpatVectors

```{r, dif}
dif <- symdif(z,p)
plot(dif, col=rainbow(length(dif)))
dif
```

## Spatial queries

Query polygons with points. 

```{r, pts}
pts <- matrix(c(6, 6.1, 5.9, 5.7, 6.4, 50, 49.9, 49.8, 49.7, 49.5), ncol=2)
spts <- vect(pts, proj4string=crs(p))
plot(z, col='light blue', lwd=2)
points(spts, col='light gray', pch=20, cex=6)
text(spts, 1:nrow(pts), col='red', font=2, cex=1.5)
lines(p, col='blue', lwd=2)
```

`extract` is used for queries between SpatVector and SpatRaster objects, and also for queries between SpatVectors.


```{r}
extract(spts, p)
extract(spts, z)
```
