# Maps

Like for other plots, there are different approaches in R to make maps. 

## Vector data

### Base plots 

```{r, maps_1}
library(terra)
p <- vect(system.file("ex/lux.shp", package="terra"))
plot(p)
```

```{r, maps_2}
n <- size(p)
plot(p, col=rainbow(n))
```

One colour per region (NAME_1)

```{r, maps_3}
u <- unique(p$NAME_1)
u
m <- match(p$NAME_1, u)
plot(p, col=rainbow(n)[m])
#text(p, 'NAME_2', cex=.75, halo=TRUE)
```


## Raster

Example data

```{r}
library(terra)
b <- rast(system.file("ex/logo.tif", package="terra"))

r <- rast(p, res=0.01 )
values(r) <- 1:ncell(r)
r <- mask(r, p)
```

Several generic functions have been implemented for SpatRaster objects to create maps and other plot types. Use 'plot' to create a map of a SpatRaster. You can zoom in using 'zoom' and clicking on the map twice (to indicate where to zoom to). With `click` it is possible to interactively query a SpatRaster by clicking once or several times on a map plot.

After plotting a `SpatRaster` you can add vector type spatial data (points, lines, polygons). You can do this with functions points, lines, polygons if you are using the basic R data structures or plot(object, add=TRUE) if you are using Spatial* objects as defined in the sp package. When plot is used with a multi-layer object, all layers are plotted (up to 16), unless the layers desired are indicated with an additional argument. 

```{r, maps_5}
plot(r)
plot(p, add=TRUE)
```

`image` does not provide a legend and that can be advantageous in some cases.

```{r, maps_6}
image(r)
plot(p, add=TRUE)
```

Multi-layer Raster objects can be plotted as individual layers
```{r, raster-20a, fig.width=7}
plot(b)
```

They can also be combined into a single image, by assigning individual layers to one of the three color channels (red, green and blue):
```{r, raster-20b }
plotRGB(b, r=1, g=2, b=3)
```

You can also use a number of other plotting functions with `SpatRasters`, including `hist`, `persp`, `contour`}, and `density`. See the help files for more info. 


