---
output: html_document
editor_options: 
  chunk_output_type: console
---

# Maps

You can make a map with `plot(x)`, were `x` is a `SpatRaster` or a `SpatVector`. You can add additional spatial data or text with functions such as `points`, `lines`, `text`

You can zoom in using `zoom(x)` and clicking on the map twice (to indicate where to zoom to). Or use `sel(x)` to save a spatial subset to a new object. With `click(x)` it is possible to interactively query a SpatRaster by clicking once or several times on a map plot.

## SpatVector

Example data 

```{r, vect_1}
library(terra)
p <- vect(system.file("ex/lux.shp", package="terra"))
```

If you plot a `SpatVector` without further arguments, you get black points, lines or polygons, and no legend.

```{r, maps_1, fig.height=5}
plot(p)
```

You can add colors like this

```{r, maps_2, fig.height=5}
n <- nrow(p)
plot(p, col=rainbow(n))
```

But if you want colors it is probably easiest to use an attribute. 

```{r, maps_3}
plot(p, "NAME_2", col=rainbow(25))
```


You can request maps for multiple variables
```{r, maps_4a, fig.width=10}
plot(p, c("NAME_1", "NAME_2"), col=rainbow(25))
```

Below we also make two maps, but do it "by hand". We adjust the spacing, and put the legends inside the map area, and use non-rotated text for the vertical axis.

```{r, maps_4b, fig.width=10}
par(mfrow=c(1,2))
m <- c(3.1, 3.1, 2.1, 2.1)
plot(p, "NAME_1", col=rainbow(25), mar=m, plg=list(x="topright"), pax=list(las=1))
plot(p, "NAME_2", col=rainbow(25), mar=m, plg=list(x="topright", cex=.75), pax=list(las=1))
```

More costumization. Choose the axes to draw, at a label and a box to the legend.

```{r, maps_5, fig.width=10}
par(mfrow=c(1,2))
m <- c(3.1, 3.1, 1.1, 1.1)
plot(p, "NAME_1", col=rainbow(25), mar=m, plg=list(x="topright", title="District", bty = "o"), main="", axes=FALSE)
axis(1, at=c(5,7)); axis(1)
axis(2, at=c(49,51)); axis(2, las=1)

plot(p, "NAME_2", col=rainbow(25), mar=m, plg=list(x="topright", cex=.75, title="Canton", bty = "o"), main="", axes=FALSE)
axis(1, at=c(5, 7)); axis(1)
```

We can combine multiple SpatVectors using `lines` and `points` to draw on top of what we plotted first.

```{r, maps_6}
d <- aggregate(p, "NAME_1")
plot(p, col="light blue", lty=2, border="red", lwd=2)
lines(d, lwd=5)
lines(d, col="white", lwd=1)
text(p, "NAME_2", cex=.8, halo=TRUE)
```

The `rasterVis` package provides a lot of very nice plotting options as well. 


## SpatRaster


Example data

```{r, maps_10}
f <- system.file("ex/elev.tif", package="terra")
r <- rast(f)
```

The default display of a single layer SpatRaster depends on the data type, but there will always be a legend.
```{r, maps_11}
plot(r)
```

After plotting a `SpatRaster` you can add vector type spatial data (points, lines, polygons). You can do this with functions `points`, `lines`, `polys` or `plot(object, add=TRUE)`. 

```{r, maps_12}
plot(r)

lines(p, lwd=2)
set.seed(12)
xy <- spatSample(r, 20, "random", na.rm=TRUE, xy=TRUE)
points(xy, pch=20, col="red", cex=2)
```

Or use a different legend type 

```{r, maps_13}
m <- c(3.1, 3.1, 1.1, 1.1)
plot(r, type="interval", plg=list(x="topright"), mar=m)
```

If there are only a few values, the default is to show "classes"

```{r, maps_14}
rr <- round(r/100)
plot(rr, plg=list(x="topright"), mar=m)
```


If the raster is categorical you get the category labels in the legend.

Make a categorical (factor) raster 

```{r, maps_15, fig.height=5, fig.width=9}
x <- classify(r, c(140, 300, 400, 550))
levels(x) <- data.frame(id=0:2, elevation=c("low", "intermediate", "high"))
is.factor(x)
x

plot(x, col=c("green", "blue", "light gray"))
```

When plot is used with a multi-layer object, all layers are plotted (up to 16), unless the layers desired are indicated with an additional argument. 


```{r, maps_16}
library(terra)
b <- rast(system.file("ex/logo.tif", package="terra"))
plot(b)
```


```{r, maps_17}
r <- rast(p, res=0.01 )
values(r) <- 1:ncell(r)
r <- mask(r, p)
```

In this case, it makes sense to combine the three layers into a single image, by assigning individual layers to one of the three color channels (red, green and blue):
```{r, maps_20}
plotRGB(b, r=1, g=2, b=3)
```

You can also use a number of other plotting functions with SpatRasters, including `hist`, `persp`, `contour`}, and `density`. See the help files for more info. 

The `rasterVis` and `tmap` packages provides a lot of very nice mapping options as well. 

## Basemaps

You can get many different base-maps with the maptiles package. Reading the data again. 
```{r, maps_data}
library(terra)
f <- system.file("ex/lux.shp", package="terra")
p <- vect(f)
```

```{r, maps_21}
library(maptiles)
bg <- get_tiles(ext(p))
plotRGB(bg)
lines(p, col="blue", lwd=3)
```

## Interactive maps 

You can use the `leaflet` package to make interactive maps. 

```{r leaflet, message=FALSE}
library(leaflet)
m <- plet(p)
m
```
